The impacts of demography on linked selection in maize

### 2015-07-02 manuscript.tex
edits/comments

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{fig:purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. \label{fig:purify} \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\end{figure*}

@@

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained purely by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\textcolor{red}{(reference here)} was therefore shifted
to more positive values \textcolor{red}{(figure here)}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. Therefore, for demographic modeling we restricted analysis
to non-genic sites.

@@

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) of
0.014734, which is similar to previous estimates
\textcolor{red}{(references here)}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \textcolor{red}{(reference here)}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{fig:bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data which suggests maize domestication began
approximately 9,000 years before present \textcolor{red}{(reference
  here)}. This could result from multiple generations per year, or in
may reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate.

@@

\section{Discussion}
\subsection{Discussion 1}
Text goes here

>>>>

{\bf Outline}
\begin{enumerate}
\item Diversity differs between genes, not genes
\item No hard sweeps
\item soft sweeps? (putting it here means if we find a lot we may have to redo stuff below on the subset of genes w/o evidence of soft sweeps.)
\item purifying selection differs ($\pi$)
\item Estimating the bneck
\item purifying selection (singletons)
\item fit the B curve
\item simulations
\end{enumerate}

\subsection{Patterns of purifying and background selection \jri{we will need to define background selection and be careful about ``purifying selection'' vs ``background selection'', as technically background selection does not include very weak deleterious alleles.}are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious mutations arising in a population are continuously selected against. 
When this form of selection is operating, it can serve to reduce genetic variability at linked neutral sites, a phenomenon called background selection \cite{charlesworth1993}. Purifying and background selection lead to lower diversity within genes and other functional sites relative to neutral regions \textcolor{red}{(ref here)}. 
We investigated purifying selection in maize and teosinte by evaluating the average magnitude of reduced diversity within genes and recovery away from genes in both taxa. 
When standardized by neutral levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity and slower recovery was observed for teosinte than for maize, implying that purifying selection has left a more pronounced signature in the teosinte genome (Figure \ref{fig:purify}). 
This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may \jri{don't think we should say this; this wasn't my a priori prediction}
have elevated the intensity of purifying selection for maize. 
We therefore conducted a parallel analysis based on singleton diversity. \jri{disagree singletons should go here. see outline above.}
As a class, singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. \label{fig:purify} \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.'' }
\end{figure*}

@@

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained purely by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\textcolor{red}{(reference here)} was therefore shifted
to more positive values \textcolor{red}{(figure here)}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. \jri{need to cite Hufford 2012 here as we made this observation in that paper} Therefore, for demographic modeling we restricted analysis
to non-genic sites. \jri{need to clarify 5kb from genes}

@@

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) of
0.014734, which is similar to previous estimates
\textcolor{red}{(references here)}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \textcolor{red}{(reference here)}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{fig:bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data which suggests maize domestication began
approximately 9,000 years before present \textcolor{red}{(reference
  here)}. This could result from multiple generations per year \jri{this seems unlikely, i'd probably drop multiple generations as an explanation}, or in
may reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate.

@@

\section{Discussion}
\subsection{Discussion 1}
Text goes here
\jri{include back of envelope calculatio based on land area!}
